---
title: "P8105 HW3"
author: "Abhishek Ajay (aa4266)"
date: "October 14, 2018"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      fig.width = 8.5,
                      fig.asp = 0.9,
                      out.width = "90%"
                     )
library(p8105.datasets)
library(tidyverse)
library(ggplot2)
theme_set(theme_bw() + theme(legend.position = "bottom"))
```

#Problem 1

Data import and cleaning

```{r p1_data_import}
brfss_data_tidy = 
  p8105.datasets::brfss_smart2010 %>% 
  janitor::clean_names() %>% 
  rename(resp_id = respid, location_abbr = locationabbr, location_desc = locationdesc) %>% #further manual renaming to appropriate variable names
  filter(topic == "Overall Health") %>% 
  select(year, location_abbr, location_desc, response, data_value)

brfss_data_tidy$response = 
  brfss_data_tidy$response %>%  
  factor(levels = c("Excellent", "Very good", "Good", "Fair", "Poor"))

brfss_data_tidy
```
Here we cleaned the data by focussing only on the "Overall  Health" topic. Under this topic the response are exhaustively "Excellent", "Very good", "Good", "Fair" and "Poor". 

We recoded the reponse column into a factor data type and set the levels from the "Excellent" to "Poor".

Now we do some *exploratory analysis* of the data in the following:

1. In 2002, which states were observed at 7 locations?

```{r p1.1_brfss}
brfss_data_tidy %>% 
  ungroup() %>% 
  filter(year == 2002) %>% 
  distinct(location_abbr, location_desc) %>% 
  group_by(location_abbr) %>% 
  count() %>% 
  filter(n == 7) %>% 
  select(location_abbr) %>% 
  knitr::kable()
```

In 2002, the states: CT, FL and NC were observed at 7 locations.

2. Make a “spaghetti plot” that shows the number of locations in each state from 2002 to 2010.

```{r p1.2_brfss}
#Restructuring the data for the plot required and Spaghetting plot using ggplot
brfss_data_tidy %>% 
  ungroup() %>% 
  group_by(year, location_abbr) %>% 
  distinct(location_abbr, location_desc) %>% 
  summarise(location_count = n()) %>% 
  ggplot(aes(x = year, y = location_count, color = location_abbr)) +
    geom_line() +
    labs(
      title = "Number of locations in each state",
      x = "Year",
      y = "Number of locations",
      color = 'States'
    ) 
```
